A formal treatment of bidirectional typing

### 2023-10-16 tex/ESOP24/S1-intro.tex
[ tex ] Update future work and accordingly S1

To give a bit more detail:
The type synthesis problem is not just about deciding whether a raw term is typable---there is a third possibility that the term does not have sufficient annotations.
To solve the type synthesis problem, before attempting to decide typability (using a bidirectional type synthesiser), we should first decide whether the raw term has sufficient annotations, which corresponds to whether the term has a mode derivation.
Our theory gives a (proof-relevant) \emph{mode decorator}, which constructs a mode derivation for a raw term or provide information that refutes the existence of any mode derivation and helps the user to pinpoint missing annotations.
Then a bidirectional type synthesiser is required to decide the typability of only mode-decorated raw terms instead of all raw terms.
Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between bidirectional typing derivations and pairs of a typing derivation and a mode derivation for the same raw term; this reformulation is more natural and useful than \varcitet[Section~3.2]{Dunfield2021}{'s annotatability}.

@@

To make our exposition accessible, the theory in this paper focuses on simply typed languages with a syntax-directed bidirectional type system, so that the decidability of bidirectional type synthesis can be established without any other technical assumptions.
%the idea of extending a signature with bidirectionality should be clear enough for further generalisation.
By adding assumptions, it should be possible to extend the theory to deal with more expressive type systems; for instance, we briefly discuss how the theory can handle polymorphically typed languages such as \SystemF with additional assumptions in \cref{sec:future}.

>>>>

To give a bit more detail:
The type synthesis problem is not just about deciding whether a raw term is typable---there is a third possibility that the term does not have sufficient annotations.
To solve the type synthesis problem, before attempting to decide typability (using a bidirectional type synthesiser), we should first decide whether the raw term has sufficient annotations, which corresponds to whether the term has a mode derivation.
Our theory gives a (proof-relevant) \emph{mode decorator}, which constructs a mode derivation for a raw term or provide information that refutes the existence of any mode derivation and helps the user to pinpoint missing annotations.
Then a bidirectional type synthesiser is required to decide the typability of only mode-decorated raw terms instead of all raw terms.
Soundness and completeness of bidirectional typing is reformulated as a one-to-one correspondence between bidirectional typing derivations and pairs of a typing derivation and a mode derivation for the same raw term; this reformulation is more natural and useful than annotatability~\cite[Section~3.2]{Dunfield2021} which is typically formulated in the literature of bidirectional typing.

@@

To make our exposition accessible, the theory in this paper focuses on simply typed languages with a syntax-directed bidirectional type system, so that the decidability of bidirectional type synthesis can be established without any other technical assumptions.
%the idea of extending a signature with bidirectionality should be clear enough for further generalisation.
By adding assumptions, it should be possible to extend the theory to deal with more expressive type systems; for instance, we briefly discuss how the theory can handle polymorphically typed languages such as \SystemF and \SystemFsub with additional assumptions in \cref{sec:future}.

### 2023-10-16 tex/ESOP24/S8-future.tex
[ tex ] Update future work and accordingly S1

\section{Discussion} \label{sec:future}
We believe that our formal treatment lays a foundation for further investigation, as the essential aspects of bidirectional typing have been studied rigorously. 
While our current development is based on simply typed languages to highlight the core ideas, it is evident that many concepts and extensions remain unexplored.
We will discuss related work and outline potential directions for future research.
\subsubsection{Language formalisation frameworks}
Many general definitions of logics and type theories and frameworks for defining them have been proposed.

@@

\varcitet{Felicissimo2023}{'s computation LF} is capable of defining dependent type theories with computational rules, similar to Dedukti, and enabling generic bidirectional type synthesis.
Generic bidirectional type synthesis, as presented in Felicissimo's work, is decidable, sound, and complete with respect to decorated terms, provided that the underlying language specification is mode-correct, and the equality check is well-behaved, i.e.\ type-preserving, confluent, and strongly normalising.
However, the last assumption is challenging to establish, even for specific languages and lacks a comprehensive general understanding.

@@

Extending general definitions for bidirectional typing and mode derivations from \citeauthor{Hamana2011}'s work is straightforward. 
For example, the universal type $\forall \alpha.\, A$ and type abstraction in \SystemF can be specified as operations:
\[
\mathsf{all} : * \rhd [*] * \to *
\quad\text{and}\quad
\mathsf{tabs} : [*] A \rhd \left< * \right> A^{\chk} \to \tyOp_{\mathsf{all}}(\alpha. A)^{\chk}
\]
The decidability of bidirectional type synthesis checking (\cref{thm:bidirectional-type-synthesis-checking}) should also carry over, as no equations are imposed on types and no guessing is required.

However, explicit type application is impractical but implicit application results in the \emph{instantiation problem} which amounts to replacing type equality checks $A = B$ with checking subtyping $A <: B$ and guessing $B$ in $\forall \alpha. A <: A[B/\alpha]$ in the presence of universal quantification.
A general theory for bidirectional typing with polymorphism and subtyping that accommodates various solutions to the instantiation problem is still under development.

>>>>

\section{Discussion} \label{sec:future}
We believe that our formal treatment lays a foundation for further investigation, as the essential aspects of bidirectional typing have been studied rigorously. 
While our current development is based on simply typed languages to highlight the core ideas, it is evident that many concepts and extensions remain unexplored.
In this section we discuss related work and outline potential directions for future research.
\subsubsection{Language formalisation frameworks}
Many general definitions of logics and type theories and frameworks for defining them have been proposed.

@@

\varcitet{Felicissimo2023}{'s computation LF} is capable of defining dependent type theories with computational rules, similar to \Dedukti, and enabling generic bidirectional type synthesis.
Generic bidirectional type synthesis, as presented in Felicissimo's work, is decidable, sound, and complete with respect to decorated terms, provided that the underlying language specification is mode-correct, and the equality check is well-behaved, i.e.\ type-preserving, confluent, and strongly normalising.
However, the last assumption is challenging to establish, even for specific languages and lacks a comprehensive general understanding.

@@

Extending general definitions for bidirectional typing and mode derivations from \citeauthor{Hamana2011}'s work is straightforward. 
For example, the universal type $\forall \alpha.\, A$ and type abstraction in \SystemF can be specified as operations:
\[
\mathsf{all} : * \rhd [*] * \to *
\quad\text{and}\quad
\mathsf{tabs} : [*] A \rhd \left< * \right> A^{\chk} \to \tyOp_{\mathsf{all}}(\alpha. A)^{\chk}
\]
The decidability of bidirectional type synthesis (\cref{thm:bidirectional-type-synthesis-checking}) should also carry over, as no equations are imposed on types and no guessing (for type application) is required.
Adding subtyping $A \mathrel{<:} B$ to languages can be done by replacing type equality with a subtyping relation $\mathrel{<:}$ and type equality check with subtyping check, so polymorphically typed languages with subtyping such as \SystemFsub can be specified.
The main idea of bidirectional typing does not change, so the formal development should extend without further assumptions as well.

However, explicit type application is impractical but its implicit counterpart results in the \emph{instantiation problem} which amounts to transforming the implicit type application (a stationary rule) to guessing $B$ in $\forall \alpha. A <: A[B/\alpha]$.
A general theory for bidirectional typing that accommodates various solutions to the instantiation problem as an assumption is still under development.
